Local Spectral Clustering of Density Upper Level Sets

### 2019-06-20 neurips/local_spectral_density_clustering_arxiv.tex
more random edits
edits to Sec 1 and 2

\begin{abstract}
\noindent We analyze Personalized PageRank (PPR), a local spectral method for clustering,
which extracts clusters using locally-biased random walks around a user-specified
seed node.  In contrast to previous work, we adopt a traditional statistical
learning setup, where we obtain samples from an unknown distribution, and aim to
identify connected regions of high-density (density clusters).  We prove that
PPR, run on a neighborhood graph, extracts sufficiently salient density
clusters. We also provide empirical support for our theory.
\end{abstract}
\end{center}
\section{Introduction}
\label{sec: introduction}

In this paper, we study the problem of clustering: splitting a given data set
into groups that satisfy some notion of within-group similarity and
between-group difference.  We focus on spectral clustering, a family of powerful
nonparametric clustering algorithms.  Generally speaking, a spectral algorithm
first constructs a geometric graph $G$, where vertices correspond to samples,
and edges correspond to proximities between samples. It then estimates a feature
embedding based on the Laplacian of $G$, and applies a simple clustering
technique (like k-means clustering) in the embedded feature space.

When applied to geometric graphs built from a large number of samples,
global spectral clustering methods can be computationally cumbersome and   
insensitive to the local geometry of the underlying distribution
\citep{leskovec2010,mahoney2012}.  This has led to increased interest in
local spectral clustering algorithms, which leverage locally-biased spectra
computed using random walks around some user-specified seed node.  A popular 
local clustering algorithm is Personalized PageRank (PPR), first introduced by 
\citet{haveliwala2003}, then further developed by
several others~\citep{spielman2011,spielman2014,andersen2006,mahoney2012,zhu2013}.

Local spectral clustering techniques have been practically very successful
\citep{leskovec2010,andersen2012,gleich2012,mahoney2012,wu2012}, leading 
many authors to develop supporting theory
\citep{spielman2013,andersen2009,gharan2012,zhu2013} that gives worst-case
guarantees on traditional graph-theoretic notions of cluster quality (such as
conductance).  In this paper, we adopt a more traditional statistical viewpoint,
and examine what the output of local clustering on a data set reveals about the
underlying density of the samples 
$f$.  In particular, we examine the ability of PPR to recover
\emph{density clusters} of $f$, defined as the connected components of
the upper level set $\{x \in \Rd : f(x) \geq \lambda\}$ for some $\lambda > 0$
(a central object of interest in the statistical clustering literature, dating
back to the work of \citet{hartigan1981}).

\paragraph{PPR on a neighborhood graph.} We now describe the clustering
algorithm that will be our focus for the rest of the paper. Let $\Xbf = \{x_1,
\ldots, x_n\}$ be a sample drawn i.i.d.\ from a distribution $\Pbb$ on $\Rd$,
with density $f$.  For a radius $r > 0$, we define $G_{n,r}=(V,E)$ to be the
\emph{$r$-neighborhood graph} of $\Xbf$, an unweighted, undirected graph with
vertices $V=\Xbf$, and an edge $(x_i,x_j) \in E$ if and only if $\norm{x_i -
x_j} \leq r$, where $\norm{\cdot}$ is the $\ell_2$ norm. We denote by $\Abf \in
\Reals^{n \times n}$ the adjacency matrix, with entries $\Abf_{uv} = 1$ if
$(u,v) \in E$ and $0$ otherwise.  We also denote by $\Dbf$ the diagonal degree
matrix, with $\Dbf_{uu} = \sum_{v \in V} \Abf_{uv}$, and by $\Ibf{}$ the $n
\times n$ identity matrix.

Next, we define the PPR vector $\pbf = \pbf(v,\alpha;G_{n,r})$, based on
a seed node $v \in V$ and a teleportation parameter $\alpha \in [0,1]$, to be 
the solution of the following linear system:
\begin{equation}
\label{eqn: ppr_vector}
\pbf = \alpha \ebf{v} + (1 - \alpha) \pbf \Wbf,
\end{equation}
where $\Wbf = (\Ibf{} + \Dbf^{-1}\Abf)/2$ is the lazy random walk matrix over
$G_{n,r}$ and $e_{v}$ is the indicator vector for node $v$ (that has a 1 in the
$v$th position and 0 elsewhere). For a level $\beta > 0$ and a target volume
$\vol_0 > 0$, we define a \emph{$\beta$-sweep cut} of $\pbf = (p_u)_{u \in V}$
as   
\begin{equation}
\label{eqn: sweep_cuts}
S_\beta := \set{u \in V: \frac{p_u}{\Dbf_{uu}} > \frac{\beta}{\vol_{0}}}.
\end{equation}
We will use the normalized cut metric to determine which sweep cut $S_{\beta}$
is the best cluster estimate. For a set $S \subseteq V$ with complement $S^c = V
\setminus S$, we define \smash{$\cut(S;G_{n,r}) := \sum_{u \in S, v \in S^c}
\Abf_{uv}$}, and \smash{$\vol(S; G_{n,r}) := \sum_{u \in S} \Dbf_{uu}$}.  We
define the \emph{normalized cut} of $S$ as
\begin{equation}
\label{eqn: normalized_cut}
\Phi(S; G_{n,r}) := \frac{\cut(S;G_{n,r})}{\min \set{\vol(S; G_{n,r}), \vol(S^c; G_{n,r})}}.
\end{equation}
Having computed sweep cuts $S_{\beta}$ over a range \smash{$\beta \in  
(\frac{1}{40},\frac{1}{11})$}\footnote{The choice of a specific range such as  
\smash{$(\frac{1}{40}, \frac{1}{11})$} is standard in the analysis of PPR
algorithms, see, e.g., \citep{zhu2013}.}, we output the cluster estimate
\smash{$\Cest = S_{\beta^*}$} that has minimum normalized cut.
%\smash{$\Phi(S_{\beta^*}; G_{n,r})$}. 
For concreteness, this is summarized in Algorithm \ref{alg: ppr}.

\begin{algorithm}
\caption{PPR on a neighborhood graph}
\label{alg: ppr}	
{\bfseries Input:} data $\Xbf=\{x_1,\ldots,x_n\}$, radius $r > 0$, teleportation
parameter $\alpha \in [0,1]$, seed $v \in \Xbf$, target stationary volume
$\vol_0 > 0$. \\     
{\bfseries Output:} cluster $\Cest \subseteq V$.
\begin{algorithmic}[1]
  \STATE Form the neighborhood graph $G_{n,r}$.
  \STATE Compute the PPR vector $p=\pbf(v, \alpha; G_{n,r})$ as in \eqref{eqn: 
    ppr_vector}. 
  \STATE For \smash{$\beta \in (\frac{1}{40}, \frac{1}{11})$} compute sweep cuts 
  $S_{\beta}$ as in \eqref{eqn: sweep_cuts}.
  \STATE Return as a cluster \smash{$\Cest = S_{\beta^*}$}, where  
  $$
  \beta^* = \argmin_{\beta \in (\frac{1}{40}, \frac{1}{11})} \Phi(S_{\beta}; G_{n,r}).
  $$
\end{algorithmic}
\end{algorithm}

\paragraph{Estimation of density clusters.} Let \smash{$\Cbb_f(\lambda)$} denote 
the connected components of the density upper level set $\{x \in \Rd: f(x) >
\lambda\}$.  For a given density cluster \smash{$\Cset \in \Cbb_f(\lambda)$}, we
call $\Cset[\Xbf] = \Cset \cap \Xbf$ the \emph{empirical density cluster}. The
size of the symmetric set difference between estimated and empirical clusters is 
a commonly used metric to quantify cluster estimation error
\citep{korostelev1993,polonik1995,rigollet2009}.

@@

\begin{definition}[Consistent density cluster estimation]
  \label{def: consistent_density_cluster_estimation}
  For an estimator \smash{$\Cest \subseteq \Xbf$} and cluster 
  \smash{$\Cset \in \Cbb_f(\lambda)$}, we say \smash{$\Cest$} is a consistent 
  estimator of $\Cset$ if for all \smash{$\Cset' \in \Cbb_f(\lambda)$} with
  $\Cset \not= \Cset'$, the following holds as $n \to \infty$: 
  \begin{equation}
    \label{eqn: consistent_density_cluster_recovery}
    \Cset[\Xbf] \subseteq \Cest \quad \text{and} \quad
    \Cest \cap \Cset'[\Xbf] = \emptyset,
  \end{equation}
  with probability tending to 1.
\end{definition}

\paragraph{Summary of results.} A summary of our results (and an outline for this
paper) is as follows.

@@

Our main takeaway: PPR, run on a neighborhood graph, recovers geometrically
compact high-density clusters.

\paragraph{Related work.} In addition to the background on local spectral
clustering given previously, a few related lines of work are worth
highlighting. \citet{shi2009} and \citet{schiebinger2015} examine the consistency of  
spectral algorithms in recovering the latent labels in certain
nonparametric mixture models. Their results focus on global rather than local 
methods, and thus impose global rather than local conditions on the nature
of the density. Moreover, they do not in general guarantee recovery of density 
clusters, which is the focus in our work. Perhaps most importantly, these works
rely on general cluster saliency conditions, which implicitly depend on many
distinct geometric aspects of the cluster $\Cset$ under consideration. We make
this dependence more explicit, and in doing so expose the role each geometric
condition plays in the clustering problem.

More broadly, density clustering and level set estimation is a well-studied
problem. \citet{polonik1995, rigollet2009} study density clustering under the 
symmetric set difference metric, \citet{tsybakov1997, singh2009} describe
minimax optimal level-set estimators under Hausdorff loss and
\citet{hartigan1981, chaudhuri2010, balakrishnan2013} consider consistent estimation of the
cluster tree.
%, to note but a few works. 
Our goal is not to improve on these
results, nor to offer a better algorithm for level set estimation; indeed, seen as
a density clustering algorithm, PPR has none of the optimality guarantees 
found in the aforementioned works. Instead, our motivation is to start with a 
widely-used local spectral method, PPR, and to better understand and
characterize the distinctions between those density clusters which are
well-conditioned for PPR, and those which are not.

@@

For arbitrary starting distribution $q$ (meaning $\textrm{supp}(q) \subseteq V$ and $\sum_{u \in V}q(u) = 1$), and for $t \geq 0$ an integer, let $q^{(t)}$ be the $t$-step probability distribution of the lazy random walk with starting distribution $q$. Consider the distance function $h_q^{(t)}, t \geq 0$, 
\begin{equation*}
h_q^{(t)}(x) = \max \left\{ \sum_{u \in V} \left(q^{(m)}(u) - \pi(u)\right)w(u)\right\}
\end{equation*}
where the maximum is over all $w: V \to [0,1]$ such that $0 \leq w(u) \leq 1$ for all $u$, and $\sum_{u \in V} w(u) \pi(u) = x$. Writing $h_v^{(t)}(x) := h_{e_v}^{(t)}(x)$ in a small abuse of notation, in Lemmas \ref{lem: lt_ub} and \ref{lem: mixing_time_small_sets}, we give an upper bound on $h_{v}^{(t)}(x)$ for all $0 \leq x \leq 1$.

>>>>

\begin{abstract}
\noindent We analyze the Personalized PageRank (PPR) algorithm, a local spectral method for clustering,
which extracts clusters using locally-biased random walks around a user-specified
seed node.  In contrast to previous work, we adopt a traditional statistical
learning setup, where we obtain samples from an unknown distribution, and aim to
identify connected regions of high-density (density clusters).  We prove that
PPR, run on a neighborhood graph, extracts sufficiently salient density
clusters. We also provide empirical support for our theory.
\end{abstract}
\end{center}
\section{Introduction}
\label{sec: introduction}
In this paper, we consider the problem of clustering: splitting a given data set
into groups that satisfy some notion of within-group similarity and
between-group difference.  Our particular focus is on
spectral clustering methods, a family of powerful
nonparametric clustering algorithms. Roughly speaking, a spectral algorithm
first constructs a geometric graph $G$, where vertices correspond to samples,
and edges correspond to proximities between samples. The algorithm 
then estimates a feature
embedding based on (an appropriate) Laplacian matrix of the graph 
$G$, and applies a simple clustering
technique (like $k$-means clustering) in the embedded feature space.

% Spectral clustering and density clustering
%Spectral clustering algorithms, which
%originated in the early work of Fiedler \cite{fiedler73}, have seen a recent revival 
%(see for instance the recent review~\cite{XX}) focused mainly on understanding the performance of spectral clustering on the Stochastic Block Model (SBM) and closely related problems.
%A parallel line of work \cite{shi09,schiebinger15}, has considered spectral partitioning applied to a non-parametric mixture model
%where once again the clustering goal is to recover the latent labels of a sample $X_1,\ldots,X_n \sim P$. These works proceed by relating the eigenvectors of the graph Laplacian to eigenfunctions of an appropriately defined integral operator, and then using assumptions about the (non-)overlap of the mixture components to show that these eigenfunctions elucidate the latent labels.

When applied to geometric graphs built from a large number of samples,
global spectral clustering methods can be computationally cumbersome and   
insensitive to the local geometry of the underlying distribution
\citep{leskovec2010,mahoney2012}.  This has led to increased interest in
local spectral clustering algorithms, which leverage locally-biased spectra
computed using random walks around some user-specified seed node.  A popular 
local clustering algorithm is the Personalized PageRank (PPR) algorithm, first introduced by 
\citet{haveliwala2003}, then further developed by
several others~\citep{spielman2011,spielman2014,andersen2006,mahoney2012,zhu2013}.

Local spectral clustering techniques have been practically very successful
\citep{leskovec2010,andersen2012,gleich2012,mahoney2012,wu2012}, leading 
many authors to develop supporting theory
\citep{spielman2013,andersen2009,gharan2012,zhu2013} that gives worst-case
guarantees on traditional graph-theoretic notions of cluster quality (such as
conductance).  In this paper, we adopt a classical statistical viewpoint,
and examine what the output of local clustering on a data set reveals about the
underlying density of the samples 
$f$.  In particular, we examine the ability of PPR to recover
\emph{density clusters} of $f$, defined as the connected components of
the upper level set $\{x \in \Rd : f(x) \geq \lambda\}$ for some $\lambda > 0$
(a central object of interest in the statistical clustering literature, dating
back to the work of \citet{hartigan1981}).

\section{Background and Related Work}
We begin by providing some standard background on the PPR algorithm
and the density clustering setup, before turning our attention to related work
and a
detailed summary of our contributions.

\subsection{PPR on a neighborhood graph} 
We now describe the clustering
algorithm that will be our focus for the rest of the paper. Let $\Xbf = \{x_1,
\ldots, x_n\}$ be a sample drawn i.i.d.\ from a distribution $\Pbb$ on $\Rd$,
with density~$f$.  For a radius $r > 0$, we define $G_{n,r}=(V,E)$ to be the
\emph{$r$-neighborhood graph} of $\Xbf$, an unweighted, undirected graph with
vertices $V=\Xbf$, and an edge $(x_i,x_j) \in E$ if and only if $\norm{x_i -
x_j} \leq r$, where $\norm{\cdot}$ is the $\ell_2$ norm. We denote by $\Abf \in
\Reals^{n \times n}$ the adjacency matrix, with entries $\Abf_{uv} = 1$ if
$(u,v) \in E$ and $0$ otherwise.  We also denote by $\Dbf$ the diagonal degree
matrix, with $\Dbf_{uu} = \sum_{v \in V} \Abf_{uv}$, and by $\Ibf{}$ the $n
\times n$ identity matrix.

Next, we define the PPR vector $\pbf = \pbf(v,\alpha;G_{n,r})$, based on
a seed node $v \in V$ and a teleportation parameter $\alpha \in [0,1]$, to be 
the solution of the following linear system:
\begin{equation}
\label{eqn: ppr_vector}
\pbf = \alpha \ebf{v} + (1 - \alpha) \pbf \Wbf,
\end{equation}
where $\Wbf = (\Ibf{} + \Dbf^{-1}\Abf)/2$ is the lazy random walk matrix over
$G_{n,r}$ and $e_{v}$ is the indicator vector for node $v$ (that has a 1 in the
$v$-th position and 0 elsewhere). For a level $\beta > 0$ and a target volume
$\vol_0 > 0$, we define a \emph{$\beta$-sweep cut} of $\pbf = (p_u)_{u \in V}$
as   
\begin{equation}
\label{eqn: sweep_cuts}
S_\beta := \set{u \in V: \frac{p_u}{\Dbf_{uu}} > \frac{\beta}{\vol_{0}}}.
\end{equation}
We will use the normalized cut metric to determine which sweep cut $S_{\beta}$
is the best cluster estimate. For a set $S \subseteq V$ with complement $S^c = V
\setminus S$, we define \smash{$\cut(S;G_{n,r}) := \sum_{u \in S, v \in S^c}
\Abf_{uv}$}, and \smash{$\vol(S; G_{n,r}) := \sum_{u \in S} \Dbf_{uu}$}.  We
define the \emph{normalized cut} of $S$ as
\begin{equation}
\label{eqn: normalized_cut}
\Phi(S; G_{n,r}) := \frac{\cut(S;G_{n,r})}{\min \set{\vol(S; G_{n,r}), \vol(S^c; G_{n,r})}}.
\end{equation}
Having computed sweep cuts $S_{\beta}$ over a range \smash{$\beta \in  
(\frac{1}{40},\frac{1}{11})$}\footnote{The choice of a specific range such as  
\smash{$(\frac{1}{40}, \frac{1}{11})$} is standard in the analysis of PPR
algorithms, see, e.g., \citep{zhu2013}.}, we output the cluster estimate
\smash{$\Cest = S_{\beta^*}$} that has minimum normalized cut.
%\smash{$\Phi(S_{\beta^*}; G_{n,r})$}. 
For concreteness, the PPR algorithm is summarized in Algorithm~\ref{alg: ppr}.

\begin{algorithm}
\caption{PPR on a neighborhood graph}
\label{alg: ppr}	
{\bfseries Input:} data $\Xbf=\{x_1,\ldots,x_n\}$, radius $r > 0$, teleportation
parameter $\alpha \in [0,1]$, seed $v \in \Xbf$, target stationary volume
$\vol_0 > 0$. \\     
{\bfseries Output:} cluster $\Cest \subseteq V$.
\begin{algorithmic}[1]
  \STATE Form the neighborhood graph $G_{n,r}$.
  \STATE Compute the PPR vector $p=\pbf(v, \alpha; G_{n,r})$ as in \eqref{eqn: 
    ppr_vector}. 
  \STATE For \smash{$\beta \in (\frac{1}{40}, \frac{1}{11})$} compute sweep cuts 
  $S_{\beta}$ as in \eqref{eqn: sweep_cuts}.
  \STATE Return the cluster \smash{$\Cest = S_{\beta^*}$}, where  
  $$
  \beta^* = \argmin_{\beta \in (\frac{1}{40}, \frac{1}{11})}~ \Phi(S_{\beta}; G_{n,r}).
  $$
\end{algorithmic}
\end{algorithm}

\subsection{Estimation of density clusters} 
Let \smash{$\Cbb_f(\lambda)$} denote 
the connected components of the density upper level set $\{x \in \Rd: f(x) >
\lambda\}$.  For a given density cluster \smash{$\Cset \in \Cbb_f(\lambda)$}, we
call $\Cset[\Xbf] = \Cset \cap \Xbf$ the \emph{empirical density cluster}. The
size of the symmetric set difference between estimated and empirical clusters is 
a commonly used metric to quantify cluster estimation error
\citep{korostelev1993,polonik1995,rigollet2009}.

@@

\begin{definition}[Consistent density cluster estimation]
  \label{def: consistent_density_cluster_estimation}
  For an estimator \smash{$\Cest \subseteq \Xbf$} and cluster 
  \smash{$\Cset \in \Cbb_f(\lambda)$}, we say \smash{$\Cest$} is a consistent 
  estimator of $\Cset$ if for all \smash{$\Cset' \in \Cbb_f(\lambda)$} with
  $\Cset \not= \Cset'$, the following holds as $n \to \infty$: 
  \begin{equation}
    \label{eqn: consistent_density_cluster_recovery}
    \Cset[\Xbf] \subseteq \Cest \quad \text{and} \quad
    \Cest \cap \Cset'[\Xbf] = \emptyset,
  \end{equation}
  with probability tending to 1.
\end{definition}
\noindent Consistent cluster recovery roughly ensures that, for a given $\lambda$, the estimated cluster $\Cest$ contains all
points in a true
density cluster $\Cset$, and simultaneously does not contain any points in any other density cluster $\Cset'$.

With these definitions in place, our broad goal will be to understand the extent to which the PPR algorithm is able to
recover a cluster which either guarantees low symmetric set difference to a true density cluster, or which consistently estimates
a true density cluster.

%\subsection{Our Contributions}

\subsection{Related Work and Our Contributions}
In addition to the background on local spectral
clustering given previously, a few related lines of work are worth
highlighting. \citet{shi2009} and \citet{schiebinger2015} examine the consistency of  
spectral algorithms in recovering the latent labels in certain
nonparametric mixture models. Their results focus on global rather than local 
methods, and thus impose global rather than local conditions on the nature
of the density. Moreover, they do not in general guarantee recovery of density 
clusters, which is the focus in our work. Perhaps most importantly, these works
rely on general cluster saliency conditions, which implicitly depend on many
distinct geometric aspects of the cluster $\Cset$ under consideration. We make
this dependence more explicit, and in doing so expose the role each geometric
condition plays in the clustering problem.

More broadly, density clustering and level set estimation is a well-studied
problem. \citet{polonik1995, rigollet2009} study density clustering under the 
symmetric set difference metric, \citet{tsybakov1997, singh2009} describe
minimax optimal level-set estimators under Hausdorff loss and
\citet{hartigan1981, chaudhuri2010, balakrishnan2013} consider consistent estimation of the
cluster tree.
%, to note but a few works. 
We emphasize that our goal is not to improve on these
results, nor to offer a better algorithm for level set estimation; indeed, seen as
a density clustering algorithm, PPR has none of the optimality guarantees 
found in the aforementioned works. Instead, our motivation is to start with a 
widely-used local spectral method, PPR, and to better understand and
characterize the distinctions between those density clusters which are
well-conditioned for PPR, and those which are not.

A summary of our results (and an outline for this
paper) is as follows.

@@

\noindent At this point we summarize one of our main takeaways: 
PPR, run on a neighborhood graph, recovers \emph{geometrically
compact} high-density clusters. Our main theoretical results, make this takeaway 
precise, and provide concrete ways of quantifying the geometric
compactness of a density cluster. To complement this, we also provide experimental
evidence that the PPR algorithm can fail to recover high-density clusters which are not
geometrically compact.

@@

For arbitrary starting distribution $q$ (meaning $\textrm{supp}(q) \subseteq V$ and $\sum_{u \in V}q(u) = 1$), and for $t \geq 0$ an integer, let $q^{(t)}$ be the $t$-step probability distribution of the lazy random walk with starting distribution $q$. Consider the distance function $h_q^{(t)}, t \geq 0$, 
\begin{equation*}
h_q^{(t)}(x) = \max_{w \in \mathcal{W}_x} \left\{ \sum_{u \in V} \left(q^{(m)}(u) - \pi(u)\right)w(u)\right\}
\end{equation*}
where the maximum is over:
\begin{align*}
\mathcal{W}_x = \left\{w: V \to [0,1] \Big\vert 0 \leq w(u) \leq 1~~\forall~u,~~\text{and}~~\sum_{u \in V} w(u) \pi(u) = x\right\}.
\end{align*}
%$w: V \to [0,1]$ such that $0 \leq w(u) \leq 1$ for all $u$, and $\sum_{u \in V} w(u) \pi(u) = x$. 
Writing $h_v^{(t)}(x) := h_{e_v}^{(t)}(x)$ in a small abuse of notation, in Lemmas \ref{lem: lt_ub} and \ref{lem: mixing_time_small_sets}, we give an upper bound on $h_{v}^{(t)}(x)$ for all $0 \leq x \leq 1$.
